Understanding and Verifying the Shape of Stencil Computations

### 2017-03-27 oopsla17/lattice-model.tex
theory of approximation

% Stencil macro
%
% Optional argument: s when we want the short form, empty otherwise.
%
% Mandatory arguments:
% 1st: c/b/f/p for centered, backward, forward, pointed
% 2st: dimension
% 3st: depth
% 4st: if pointed/nonpointed needs to be specified leave it blank if omitted.
%
% Example: \stencil[s]{f}{1}{2}{} = fwd(dim=1,depth=2)
% Example: \stencil{b}{1}{2}{pointed} = backward(dim=1,depth=2,pointed)
\newcommand{\stencil}[5][]{%
  \ifthenelse{\equal{#2}{f}}{%
    \ifthenelse{\equal{#1}{s}}{%
      \texttt{fwd}%
    }{\ifthenelse{\isempty{#1}}{%
        \texttt{forward}%
      }{INVALID OPTIONAL}%
    }%
  }{\ifthenelse{\equal{#2}{b}}{%
      \ifthenelse{\equal{#1}{s}}{%
          \texttt{bwd}%
      }{\ifthenelse{\isempty{#1}}{%
          \texttt{backward}%
        }{INVALID OPTIONAL}%
      }%
  }{\ifthenelse{\equal{#2}{c}}{%
      \ifthenelse{\equal{#1}{s}}{%
          \texttt{ctd}%
      }{\ifthenelse{\isempty{#1}}{%
          \texttt{centered}%
        }{INVALID OPTIONAL}%
      }%
  }{\ifthenelse{\equal{#2}{p}}{%
      \ifthenelse{\equal{#1}{s}}{%
          \texttt{ptd}%
      }{\ifthenelse{\isempty{#1}}{%
          \texttt{pointed}%
        }{INVALID OPTIONAL}%
      }%
  }{}}}}%
  \texttt{(dim=\textnormal{#3}%
    \ifthenelse{\isempty{#4}}{}{,depth=\textnormal{#4}}%
    \ifthenelse{\isempty{#5}}{}{,\textnormal{#5}}%
  )}%
}

@@

@@

\begin{description}
  \item[Basic] \texttt{*} and \texttt{+} are both idempotent, commutative, and
    associative;
%
  \item[Dist] \texttt{*} distributed over \texttt{+} and dually
    \texttt{+} distributes over \texttt{*}, meaning if \textcap{R}, \textcap{S},
    and \textcap{T} are regions, then we have the following equivalences:
%
    \begin{align*}
      \texttt{\textcap{R}*(\textcap{S}+\textcap{T})} & \equiv
        \texttt{(\textcap{R}*\textcap{S})+(\textcap{R}*\textcap{T})} &
      \texttt{\textcap{R}+(\textcap{S}*\textcap{T})} & \equiv
        \texttt{(\textcap{R}+\textcap{S})*(\textcap{R}+\textcap{T})}
    \end{align*}
%
  \item[Overlapping pointed] If \textcap{R} is one of \texttt{forward},
    \texttt{backward}, or \texttt{centered}, then we have
    $\texttt{\textcap{R}(depth=$k$,dim=$n$,\_) + \stencil[s]{p}{$n$}{}{}} \equiv
    \texttt{\textcap{R}(depth=$k$,dim=$n$)}$.
%
  \item[Centered] The region constants \texttt{forward} and \texttt{backward}
    are two halves of \texttt{centered} specifications:
%
    \begin{equation*}
      \stencil[s]{c}{$n$}{$k$}{\textcap{p1}} \equiv
        \stencil[s]{f}{$n$}{$k$}{\textcap{p2}} \texttt{+}
        \stencil[s]{f}{$n$}{$k$}{\textcap{p3}}
    \end{equation*}
%
    Here \textcap{p1} is \texttt{nonpointed} if both \textcap{p2} and
    \textcap{p3} are \texttt{nonpointed} and \texttt{pointed} otherwise.
%
  \item[Point] Let \textcap{R} and \textcap{S} be region constants with two
    variants a superscript \textit{f} to mean nonpointed and \textit{t} to mean
    pointed and \textcap{dr} \& \textcap{ds} be the dimensions \textcap{R} and
    \textcap{S} acts on. Then we have the following dual equivalences:
%
    \begin{align*}
      (\textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{f}})
      \;\texttt{+}\;
      (\textcap{R}^{\textit{f}} \texttt{*} \textcap{S}^{\textit{t}})
      \;\texttt{+}\;
      \texttt{p(dim=\textcap{dr})*p(dim=\textcap{ds})} & =
        \textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{t}} \\
%
      (\textcap{R}^{\textit{t}} \texttt{+} \textcap{S}^{\textit{f}}) \texttt{*}
      (\textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{t}}) & =
        \textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{f}}
        \;\texttt{+}\;
        \texttt{p(dim=\textcap{dr})*p(dim=\textcap{ds})}
    \end{align*}
\end{description}

\begin{thm}{(Soundness)}
  Equational theory is sound with respect to the lattice model. Let $R$
  and $S$ be $N$ dimensional regions, then we have
%
  \begin{equation*}
    R \equiv S \implies \interp{R}_N \equiv \interp{S}_N
  \end{equation*}
\end{thm}

\begin{proof}
  Both $\equiv$ and $=$ are equivalence relations. We only need to show that
  each property associated with specifications equivalence holds as an equality
  in the model.
%
  \begin{description}
    \item[Basic] These follow from basic properties of lattices.
%
    \item[Dist] This is immediate from the definition of distributive
      lattice which the model is one.
%
    \item[Overlapping pointed] Let $\interp{\textcap{n}}$ be $n$. Counterparts
      of individual regions for some Boolean $p$ are given below:
%
      \begin{align*}
        \interp{\texttt{\textcap{R}(depth=\textcap{k},dim=\textcap{n},\_)}}_N
        & = \vecgen{N}{n}{\interv{-k}{k}{p}} \\
%
        \interp{\texttt{pointed(dim=\textcap{n})}}_N
        &= \vecgen{N}{n}{\interv{0}{0}{\textsf{true}}} \\
%
        \interp{\texttt{\textcap{R}(depth=\textcap{k},dim=\textcap{n})}}_N
        &= \vecgen{N}{n}{\interv{-k}{k}{\textsf{true}}}
      \end{align*}

      Using \Cref{lem:vector-union}, it is sufficient to show
      $ \interv{-k}{k}{p} \;\cup\; \interv{0}{0}{\textsf{true}} =
        \interv{-k}{k}{\textsf{true}} $ holds. This immediately follows from
      \Cref{lem:zinf-identities}.
%
    \item[Centered] Let $\interp{\textcap{n}}$, $\interp{\textcap{k}}$,
      $\interp{\textcap{p1}}$, $\interp{\textcap{p2}}$, and
      $\interp{\textcap{p3}}$ be $n$, $k$, $p1$, $p2$, and $p3$ respectively. We
      also know that $p1$ is $p2 \vee p3$ from \textbf{Centered} detinition.
      Then we get the following interpretation for region constants involved in
      the equivalence:
%
      \begin{align*}
        \interp{\texttt{centered(depth=\textcap{k},dim=\textit{n},\textcap{p1})}}_N
          &= \vecgen{N}{n}{\interv{-k}{k}{p2 \vee p3}} \\
%
        \interp{\texttt{forward(depth=\textcap{k},dim=\textit{n},\textcap{p2})}}_N
          &= \vecgen{N}{n}{\interv{0}{k}{p2} }\\
%
        \interp{\texttt{backward(depth=\textcap{k},dim=\textit{n},\textcap{p3})}}_N
          &= \vecgen{N}{n}{\interv{0}{k}{p3}}
      \end{align*}
%
      Using \Cref{lem:vector-union}, it is sufficient to show $\interv{-k}{k}{p2
      \vee p3} = \interv{0}{k}{p2} \cup \interv{-k}{0}{p3}$ holds. This
      immediately follows from \Cref{lem:zinf-identities}.

@@

\begin{description}
  \item[Centered]
  \item[Point] Let \textcap{R} be a region constant with two variants: a
    superscript $f$ to mean nonpointed and $t$ to mean pointed. Then we have
    $\textcap{R}^t \preceq \textcap{R}^f$.
\end{description}

>>>>

% Stencil macro
%
% Optional argument: s when we want the short form, empty otherwise.
%
% Mandatory arguments:
% 1st: c/b/f/p for centered, backward, forward, pointed
% 2st: dimension
% 3st: depth
% 4st: if pointed/nonpointed needs to be specified leave it blank if omitted.
%
% Example: \stencil[s]{f}{1}{2}{} = fwd(dim=1,depth=2)
% Example: \stencil{b}{1}{2}{pointed} = backward(dim=1,depth=2,pointed)
\newcommand{\stencil}[5][]{%
  \ifthenelse{\equal{#2}{f}}{%
    \ifthenelse{\equal{#1}{s}}{%
      \texttt{fwd}%
    }{\ifthenelse{\isempty{#1}}{%
        \texttt{forward}%
      }{INVALID OPTIONAL}%
    }%
  }{\ifthenelse{\equal{#2}{b}}{%
      \ifthenelse{\equal{#1}{s}}{%
          \texttt{bwd}%
      }{\ifthenelse{\isempty{#1}}{%
          \texttt{backward}%
        }{INVALID OPTIONAL}%
      }%
  }{\ifthenelse{\equal{#2}{c}}{%
      \ifthenelse{\equal{#1}{s}}{%
          \texttt{ctd}%
      }{\ifthenelse{\isempty{#1}}{%
          \texttt{centered}%
        }{INVALID OPTIONAL}%
      }%
  }{\ifthenelse{\equal{#2}{p}}{%
      \ifthenelse{\equal{#1}{s}}{%
          \texttt{ptd}%
      }{\ifthenelse{\isempty{#1}}{%
          \texttt{pointed}%
        }{INVALID OPTIONAL}%
      }%
  }{\ifthenelse{\isempty{#1}}{%
      \textcap{#2}%
    }{NO OPTIONAL FOR CUSTOM}%
  }}}}%
  \texttt{(dim=\textnormal{#3}%
    \ifthenelse{\isempty{#4}}{}{,depth=\textnormal{#4}}%
    \ifthenelse{\isempty{#5}}{}{,\textnormal{#5}}%
  )}%
}

@@

\definecolor{version}{RGB}{163,193,173}
\newcommand{\longv}[1]{{\color{version} Long version:} #1}
\newcommand{\shortv}[1]{{\color{version} Short version:} #1}

@@

\begin{description}
  \item[Basic] \texttt{*} and \texttt{+} are both idempotent, commutative, and
    associative;
%
  \item[Subsumption] If $S$ and $R$ are regions with $S \preceq R$, then
    $S \texttt{+} R \equiv S$ and $S \texttt{*} R \equiv R$.
%
  \item[Dist] \texttt{*} distributed over \texttt{+} and dually
    \texttt{+} distributes over \texttt{*}, meaning if \textcap{R}, \textcap{S},
    and \textcap{T} are regions, then we have the following equivalences:
%
    \begin{align*}
      \texttt{\textcap{R}*(\textcap{S}+\textcap{T})} & \equiv
        \texttt{(\textcap{R}*\textcap{S})+(\textcap{R}*\textcap{T})} &
      \texttt{\textcap{R}+(\textcap{S}*\textcap{T})} & \equiv
        \texttt{(\textcap{R}+\textcap{S})*(\textcap{R}+\textcap{T})}
    \end{align*}
%
  \item[Overlapping pointed] If \textcap{R} is one of \texttt{forward},
    \texttt{backward}, or \texttt{centered}, then we have
    $\stencil{\textcap{R}}{$n$}{$k$}{\texttt{nonpointed}} \;\texttt{+}\;
     \stencil[s]{p}{$n$}{}{} \equiv
     \stencil{\textcap{R}}{$n$}{$k$}{}$.
%
  \item[Centered] The region constants \texttt{forward} and \texttt{backward}
    are two halves of \texttt{centered} specifications:
%
    \begin{equation*}
      \stencil[s]{c}{$n$}{$k$}{\textcap{p1}} \equiv
        \stencil[s]{f}{$n$}{$k$}{\textcap{p2}} \texttt{+}
        \stencil[s]{b}{$n$}{$k$}{\textcap{p3}}
    \end{equation*}
%
    Here \textcap{p1} is \texttt{nonpointed} if both \textcap{p2} and
    \textcap{p3} are \texttt{nonpointed} and \texttt{pointed} otherwise.
%
  \item[Point] Let \textcap{R} and \textcap{S} be region constants with two
    variants a superscript \textit{f} to mean nonpointed and \textit{t} to mean
    pointed and \textcap{dr} \& \textcap{ds} be the dimensions \textcap{R} and
    \textcap{S} acts on. Then we have the following dual equivalences:
%
    \begin{align*}
      (\textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{f}})
      \;\texttt{+}\;
      (\textcap{R}^{\textit{f}} \texttt{*} \textcap{S}^{\textit{t}})
      \;\texttt{+}\;
      \stencil[s]{p}{\textit{dr}}{}{}*\stencil[s]{p}{\textit{ds}}{}{} & =
        \textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{t}} \\
%
      (\textcap{R}^{\textit{t}} \texttt{+} \textcap{S}^{\textit{f}}) \texttt{*}
      (\textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{t}}) & =
        \textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{f}}
        \;\texttt{+}\;
        \stencil[s]{p}{\textit{dr}}{}{}*\stencil[s]{p}{\textit{ds}}{}{}
    \end{align*}
\end{description}

\begin{thm}{(Equational soundness)}
  Equational theory is sound with respect to the lattice model. Let $R$
  and $S$ be $N$ dimensional regions, then we have
%
  \begin{equation*}
    R \equiv S \implies \interp{R}_N \equiv \interp{S}_N
  \end{equation*}
\end{thm}

\begin{proof}
  Both $\equiv$ and $=$ are equivalence relations. We only need to show that
  each property associated with specifications equivalence holds as an equality
  in the model.
%
  \begin{description}
    \item[Basic \& Subsumption] These follow from basic properties of lattices.
%
    \item[Dist] This is immediate from the definition of distributive
      lattice which the model is one.
%
    \item[Overlapping pointed] Counterparts of individual regions for some
      Boolean $p$ are given below:
%
      \begin{align*}
        \interp{\stencil{\textcap{R}}{$n$}{$k$}{\texttt{nonpointed}}}_N
        & = \vecgen{N}{n}{\interv{-k}{k}{\textsf{false}}} \\
%
        \interp{\stencil{p}{$n$}{}{}}_N
        &= \vecgen{N}{n}{\interv{0}{0}{\textsf{true}}} \\
%
        \interp{\stencil{\textcap{R}}{$n$}{$k$}{}}_N
        &= \vecgen{N}{n}{\interv{-k}{k}{\textsf{true}}}
      \end{align*}

      Using \Cref{lem:vector-union}, it is sufficient to show
      $ \interv{-k}{k}{p} \;\cup\; \interv{0}{0}{\textsf{true}} =
        \interv{-k}{k}{\textsf{true}} $ holds. This immediately follows from
      \Cref{lem:zinf-identities}.
%
    \item[Centered] Let $\interp{\textcap{p1}}$, $\interp{\textcap{p2}}$, and
      $\interp{\textcap{p3}}$ be $p1$, $p2$, and $p3$ respectively. We
      also know that $p1$ is $p2 \vee p3$ from \textbf{Centered} detinition.
      Then we get the following interpretation for region constants involved in
      the equivalence:
%
      \begin{align*}
        \interp{\stencil{c}{$k$}{$n$}{$p1$}}_N
          &= \vecgen{N}{n}{\interv{-k}{k}{p2 \vee p3}} \\
%
        \interp{\stencil{f}{$k$}{$n$}{$p2$}}_N
          &= \vecgen{N}{n}{\interv{0}{k}{p2} }\\
%
        \interp{\stencil{b}{$k$}{$n$}{$p3$}}_N
          &= \vecgen{N}{n}{\interv{0}{k}{p3}}
      \end{align*}
%
      Using \Cref{lem:vector-union}, it is sufficient to show $\interv{-k}{k}{p2
      \vee p3} = \interv{0}{k}{p2} \cup \interv{-k}{0}{p3}$ holds. This
      immediately follows from \Cref{lem:zinf-identities}.

@@

\begin{description}
  \item[Equivalence] If $S$ and $R$ are regions and $S \equiv R$, then we have
    $S \preceq R$.
%
  \item[Combined] If $S$ and $R$ are regions, then we have
    $S \texttt{+} R \preceq S$ and $S \preceq S \texttt{*} R$.
%
  \item[Depth] Let $k$ and $l$ be in positive integers and $k \leq l$, $n$ some
    fixed dimension, and \textcap{p} either \texttt{pointed} or
    \texttt{nonpointed}. Further let \textcap{R} be one of \texttt{centered},
    \texttt{forward}, and \texttt{backward}. We then have
%
    \begin{equation*}
      \stencil{R}{$n$}{$l$}{\textcap{p}} \preceq \stencil{R}{$n$}{$k$}{\textcap{p}}
    \end{equation*}
%
\end{description}

\begin{thm}{(Approximation soundness)}
  Theory of approximation is sound with respect to the lattice model. Let $R$
  and $S$ be $N$ dimensional regions, then we have
%
  \begin{equation*}
    R \preceq S \implies \interp{R}_N \supseteq \interp{S}_N
  \end{equation*}
\end{thm}
%
\begin{proof}
  Both $\preceq$ and $\supseteq$ are partial orders on sets. We only need to
  show that each inequality holds in the model.
%
  \begin{description}
    \item[Equivalence] Region equavilance is modeled with set equality and the
      partial order on regions is modeled with superset relation. Partial orders
      are reflexive and the model of eqivalence agrees with that of the partial
      order, so this rule holds.
%
    \item[Combined] \texttt{*} and \texttt{+} maps to join and meet operations
      in the lattice. All joins and meets in a lattice satisfy these
      inequalities.

    \item[Depth] When $m$ is $k$ and $l$ in
      $\stencil{\textcap{R}}{$n$}{$m$}{\textcap{p}}$, we obtain the
      interpretations $\vecgen{N}{n}{\textit{int}_k}$ and
      $\vecgen{N}{n}{\textit{int}_l}$ for some $N$ as interpretations of
      regions. Call these \textit{ks} and \textit{ls} respectively. One way to
      show $\textit{ls} \supseteq \textit{ks}$ is to show that the intersection
      of \textit{ls} and \textit{ks} is \textit{ks}.

      To show that the intersection is \textit{ks}, we use
      \Cref{lem:vector-intersect} which reduces the problem to showing that the
      intersection of $\textit{int}_k$ and $\textit{int}_l$ is the former since
      at every point except $n$ the intervals agree, hence the intersection
      remains the same. By assumption we have $k \leq l$, so the maximum of the
      lower bounds is $-k$ and the minimum of lower bounds is $k$ this is same
      as interval $\textit{int}_k$. Note that pointedness is the same for both
      intervals and depths are positive integers, so the pointedness is
      preserved under intersection. This shows the intersection of \textit{ls}
      and \textit{ks} is indeed \textit{ks} and concludes the proof.
  \end{description}
\end{proof}

We present few inequalities that can be derived from the axioms and are useful
when writing specifications without inference:

\begin{prop}{(Centered approximation)}
  For any dimension $n$, depth $k$, and pointed attribute $p$,
  we have
%
  \begin{align*}
    \stencil{c}{$n$}{$k$}{\textcap{p}} & \preceq
      \stencil{f}{$n$}{$k$}{\textcap{p}} \\
%
    \stencil{c}{$n$}{$k$}{\textcap{p}} & \preceq
      \stencil{b}{$n$}{$k$}{\textcap{p}}
  \end{align*}
\end{prop}
%
\begin{proof}
  \shortv{%
    Both inequalities are proved by breaking \texttt{centered} using
    (\textbf{Centered}) followed by the inequality (\textbf{Combined}).
  }

  \longv{%
    Proof of the first inequality:
    \begin{align*}
      \stencil[s]{c}{$n$}{$k$}{\textcap{p}}
        & \equiv \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
                 \;\texttt{+}\;
                 \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
          & (\textbf{Centered}) \\
        & \preceq \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
          & (\textbf{Combined})
    \end{align*}

    Proof of the second inequality:
    \begin{align*}
      \stencil[s]{c}{$n$}{$k$}{\textcap{p}}
        & \equiv \stencil[s]{f}{$n$}{$k$}{\textcap{p}}
                 \;\texttt{+}\;
                 \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
          & (\textbf{Centered}) \\
        & \preceq \stencil[s]{b}{$n$}{$k$}{\textcap{p}}
          & (\textbf{Combined})
    \end{align*}
  }
\end{proof}

\begin{prop}{(Point approximation)}
  Let \textcap{R} be one of \texttt{forward}, \texttt{backward}, and
  \texttt{centered}, $n$ a fixed dimension, and $k$ a fixed depth, then we have
%
  \begin{align*}
    \stencil{R}{$n$}{$k$}{} & \preceq \stencil{p}{$n$}{}{} \\
%
    \stencil{R}{$n$}{$k$}{} & \preceq
      \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
  \end{align*}
\end{prop}
%
\begin{proof}
  \shortv{%
    Both inequalities are proved by breaking \textcap{R} using
    (\textbf{Overlapping pointed}) followed by the inequality
    (\textbf{Combined}).
  }

  \longv{%
    Proof of the first inequality:
    \begin{align*}
      \stencil{R}{$n$}{$k$}{}
        & \equiv \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
                 \;\texttt{+}\;
                 \stencil[s]{p}{$n$}{}{}
        & (\textbf{Overlapping pointed}) \\
      & \preceq \stencil[s]{p}{$n$}{}{} & (\textbf{Combined})
    \end{align*}

    Proof of the second inequality:
    \begin{align*}
      \stencil{R}{$n$}{$k$}{}
        & \equiv \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
                 \;\texttt{+}\;
                 \stencil[s]{p}{$n$}{}{}
        & (\textbf{Overlapping pointed}) \\
      & \preceq \stencil{R}{$n$}{$k$}{\texttt{nonpointed}}
        & (\textbf{Combined})
    \end{align*}
  }
\end{proof}
